\chapter{Reference: the Program concept}
\label{ch:refman-program}

\begin{aside}
The smallest contract a \maya{} app must satisfy. If your
struct passes \code{static\_assert(Program<App>)}, it will
run.
\end{aside}

\section{Required members}

A type \code{P} satisfies \code{Program} when it provides:

\begin{itemize}[leftmargin=*]
  \item a nested \code{Model} type;
  \item a nested \code{Msg} type (usually a
    \code{std::variant});
  \item \code{static Model init()};
  \item \code{static std::pair<Model, Cmd<Msg>>
    update(Model, Msg)};
  \item \code{static Element view(const Model\&)};
  \item \code{static Sub<Msg> subscribe(const Model\&)}.
\end{itemize}

\section{Minimal program}

\begin{lstlisting}[language=C++]
#include <maya/maya.hpp>

using namespace maya;
using namespace maya::dsl;

struct Hello {
    struct Model {};
    struct Quit {};
    using Msg = std::variant<Quit>;

    static Model init() { return {}; }

    static auto update(Model m, Msg)
        -> std::pair<Model, Cmd<Msg>>
    {
        return {m, Cmd<Msg>::quit()};
    }

    static Element view(const Model&) {
        return t<"press q to quit"> | Bold;
    }

    static auto subscribe(const Model&) -> Sub<Msg> {
        return key_map<Msg>({ {'q', Quit{}} });
    }
};

static_assert(Program<Hello>);

int main() {
    run<Hello>({.title = "hello"});
}
\end{lstlisting}

\section{Notes}

\begin{itemize}[leftmargin=*]
  \item \code{static\_assert(Program<Hello>)} is optional but
    cheap. It surfaces missing members at compile time
    rather than as deeply nested template errors.
  \item \code{init()} can take config args if you wrap
    \code{run<>} accordingly; the bare form takes none.
  \item \code{update} returns a \emph{pair}, not just a
    Model. The second slot is the command list.
  \item Use \code{Cmd<Msg>\{\}} for "no side effect."
\end{itemize}

\section{Common mistakes}

\begin{itemize}[leftmargin=*]
  \item Returning \code{Model} from \code{update} instead of
    a pair. The concept fails; the error is loud.
  \item Forgetting \code{Quit} in the variant. Your program
    runs but cannot exit cleanly.
  \item Putting state in static variables instead of the
    Model. Works once; breaks the moment you want to test.
\end{itemize}

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Six members; one concept.
  \item Pure functions; data effects.
  \item Verify with a static assert.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Delete \code{subscribe} from the example. What does
    the compiler say?
  \item Restore it; delete \code{init}. Same question.
  \item Add a second variant case and a key for it.
\end{enumerate}
