% main.tex --- The Maya Book
%
% Build:   make            (in this directory; uses pdflatex twice)
% Output:  maya-book.pdf
%
% Chapter sources live under chapters/. To rearrange, edit \include lines.

\documentclass[11pt,openany,oneside]{book}

\input{preamble}

\date{}

\begin{document}

\frontmatter

% ----------------------------------------------------------------------
% Cover page --- one word, one line
% ----------------------------------------------------------------------
\begin{titlepage}
\thispagestyle{empty}

\vspace*{\fill}

\begin{center}
  {\fontsize{96}{100}\selectfont\sffamily\bfseries\color{mayablue} maya}
\end{center}

\vspace{0.6cm}

\begin{center}
  {\color{mayablue}\rule{4cm}{0.6pt}}
\end{center}

\vspace{0.6cm}

\begin{center}
  {\fontsize{12}{16}\selectfont\sffamily\color{mayagrey}
    a c\nolinebreak\hspace{-.05em}\raisebox{.2ex}{\small{++}}26 terminal ui library, from the byte up\par}
\end{center}

\vspace*{\fill}

\begin{center}
  {\fontsize{9}{12}\selectfont\sffamily\color{mayagrey}
    companion to the \textbf{agentty} sources\par}
\end{center}

\vspace*{1.5cm}
\end{titlepage}

% ----------------------------------------------------------------------
% Half-title / abstract
% ----------------------------------------------------------------------
\thispagestyle{empty}
\vspace*{\fill}
\begin{center}
\begin{minipage}{0.78\textwidth}
\small\itshape
This book is a deep guide to \maya{}, the header-mostly \cpp{}26
terminal-UI library that powers \agentty. We start with the terminal
as a machine, work up through the \cpp{} language features \maya{}
relies on, and then walk every layer of the library --- the DSL,
elements, the layout engine, the SIMD-accelerated renderer, the
signal graph, the Elm-style runtime, and the widget catalogue. Each
chapter ends with a working program.

\medskip

Everything in this book is grounded in the actual source under
\file{maya/}. Where the code disagrees with the prose, the code wins
--- report the mismatch as a bug.
\end{minipage}
\end{center}
\vspace*{\fill}
\clearpage

% ----------------------------------------------------------------------
% Preface
% ----------------------------------------------------------------------
\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

\section*{Who this book is for}

You have probably opened a terminal today. You have probably also
used a program that draws into it --- \code{htop}, \code{vim}, a git
client, a music player, a chat tool. This book is for the reader who
wants to write such programs themselves, in modern \cpp{}, and who
wants to understand every layer between the keystroke and the
flickering glyph.

It assumes nothing. You do not need to have written a TUI before. You
do not need to know \cpp{}26. You do not need to know what a pty is,
or what an ANSI escape sequence is, or why anyone would put a
compile-time string inside a type. Each of those is introduced from
zero, with the smallest example that makes the idea concrete, and
then connected to the place in \maya{}'s source where you will see
it again.

If you already know \cpp{}26 fluently and just want the \maya{}
tour, you can skim Chapter~\ref{ch:cpp}. Everyone else: read in
order. The book is a rope ladder; each rung sits on the one below
it.

\section*{What you will be able to do by the end}

Four concrete things, in this order:

\begin{enumerate}[leftmargin=*]
  \item Read the bytes flowing between a program and a terminal,
    and write a TUI ``Hello, red world'' in plain \cpp{} with no
    library at all (Chapter~\ref{ch:terminal}).
  \item Read every line of every \maya{} header --- templates,
    NTTPs, concepts, variants, \code{constexpr} chains --- and know
    what each construct is doing (Chapter~\ref{ch:cpp}).
  \item Build a non-trivial \maya{} program of your own, from a
    \code{Program} struct through a \code{view} function to a
    \code{Cmd<Msg>} effect, and trust that the compile-time checks
    catch the impossible states (Parts II and IV).
  \item Open the SIMD diff loop, the signal graph, the layout
    engine, or the inline-mode witness chain and follow what each
    one is doing on its own terms (Part III).
\end{enumerate}

That is the contract. If a chapter does not move you closer to one
of those four outcomes, file an issue.

\section*{How the book is shaped}

Four parts, mirroring four levels of zoom.

\paragraph{Part I --- Foundations.} The terminal as a state machine
speaking bytes; the \cpp{} language features \maya{} relies on. The
output of Part I is a reader who can run the examples and read the
headers.

\paragraph{Part II --- Architecture.} The Elm-style runtime
(\code{init}/\code{update}/\code{view}/\code{subscribe}), the
compile-time DSL that produces UI descriptions, the in-memory
\code{Element} tree, and the flexbox layout engine that places
them. The output of Part II is a reader who can build apps.

\paragraph{Part III --- Rendering.} Cells and canvases, SIMD-accelerated
frame diffing, the ANSI byte serialiser, and \maya{}'s unusual inline
mode that lives inside your scrollback instead of taking over the
terminal. The output of Part III is a reader who understands why
\maya{} feels fast.

\paragraph{Part IV --- Reactivity, runtime, and widgets.} The signal
graph that powers fine-grained reactivity, the input parser, the
full \code{Cmd}/\code{Sub} effect algebra, the catalogue of widgets,
and how to package and ship a binary built on \maya{}. The output
of Part IV is a reader who can ship.

\section*{How to read the boxes}

Five coloured boxes recur throughout. They are signposts; you can
read the book without them, but they tell you what the prose around
them is doing.

\begin{idea}
\textbf{Idea} (blue). One sentence summarising the key intuition of
the surrounding section. If you skim, read these.
\end{idea}

\begin{theory}
\textbf{Theory} (purple). The background concept from computer
science or programming-language theory behind what we just
explained. Skippable on first read; useful on second.
\end{theory}

\begin{warning}
\textbf{Pitfall} (amber). A mistake people new to the material
commonly make, and how to avoid it.
\end{warning}

\begin{recap}
\textbf{Recap} (green). A two-or-three-sentence summary of the
section just finished.
\end{recap}

\begin{terminalbox}
\textbf{Terminal} (black). Literal bytes or output, the way the
terminal sees or shows them.
\end{terminalbox}

Code in monospace with a coloured rule on the left is \cpp{} (or
occasionally shell). Lines are numbered. Long examples are split
into pieces and each piece is decoded after.

\section*{A note on accuracy}

Every code fragment in this book is either copied from \maya{}'s
headers or paraphrased from them. Where we simplify --- to keep an
example short, or to elide an implementation detail not yet
introduced --- we say so and name the file where the real version
lives. Read the book with the source open in another window;
that is the intended posture.

\bigskip

\noindent The chapters that follow assume nothing and explain
everything. Let us begin with the terminal.

\clearpage

\tableofcontents

\mainmatter

% ---- Part I: Foundations -----------------------------------------------
\part{Foundations}
\include{chapters/01-introduction}
\include{chapters/02-terminal}
\include{chapters/03-cpp}

% ---- Part II: Architecture ---------------------------------------------
\part{Architecture}
\include{chapters/04-elm}
\include{chapters/05-dsl}
\include{chapters/06-elements}
\include{chapters/07-layout}
\include{chapters/07a-concepts}
\include{chapters/07b-expected}
\include{chapters/07c-pipeline}
\include{chapters/07d-style}
\include{chapters/07e-builders}
\include{chapters/07f-text}
\include{chapters/07g-platform}
\include{chapters/07h-focus}
\include{chapters/07i-scroll}
\include{chapters/07j-errorboundary}
\include{chapters/07k-reactivity}
\include{chapters/07l-runtime}
\include{chapters/07m-testing}
\include{chapters/07n-widgetarch}
\include{chapters/07o-patterns}
\include{chapters/07p-performance}
\include{chapters/07q-guarantees}
\include{chapters/07r-cpp26}
\include{chapters/07s-comparison}
\include{chapters/07t-endtoend}
\include{chapters/07u-dispatch}
\include{chapters/07v-constexpr}
\include{chapters/07w-identity}
\include{chapters/07x-config}
\include{chapters/07y-observability}
\include{chapters/07z-i18n}
\include{chapters/07aa-lifecycle}
\include{chapters/07bb-a11y}
\include{chapters/07cc-security}
\include{chapters/07dd-cross}
\include{chapters/07ee-devx}
\include{chapters/07ff-summary}
\include{chapters/07gg-glossary}

% ---- Part III: Rendering -----------------------------------------------
\part{Rendering}
\include{chapters/08-canvas}
\include{chapters/09-simd}
\include{chapters/10-ansi}
\include{chapters/11-inline}
\include{chapters/11a-buffers}
\include{chapters/11b-diffloop}
\include{chapters/11c-colour}
\include{chapters/11d-width}
\include{chapters/11e-scheduler}
\include{chapters/11f-borders}
\include{chapters/11g-cursor}
\include{chapters/11h-mouse}
\include{chapters/11i-keyboard}
\include{chapters/11j-altscreen}
\include{chapters/11k-images}
\include{chapters/11l-hyperlinks}
\include{chapters/11m-titles}
\include{chapters/11n-clipping}
\include{chapters/11o-zorder}
\include{chapters/11p-pipeline2}
\include{chapters/11q-shaping}
\include{chapters/11r-allocations}
\include{chapters/11s-backend}
\include{chapters/11t-tracing}
\include{chapters/11u-snapshot}
\include{chapters/11v-bench}
\include{chapters/11w-slowlinks}
\include{chapters/11x-reconcile}
\include{chapters/11y-animations}
\include{chapters/11z-sprite}
\include{chapters/11ab-golden}
\include{chapters/11ac-reflow}

% ---- Part IV: Reactivity, runtime, widgets -----------------------------
\part{Reactivity, runtime, and widgets}
\include{chapters/12-signals}
\include{chapters/12a-signalgraph}
\include{chapters/12b-owners}
\include{chapters/12c-effects}
\include{chapters/12d-clocks}
\include{chapters/12e-async}
\include{chapters/12f-workers}
\include{chapters/12g-stores}
\include{chapters/12h-routing}
\include{chapters/12i-forms}
\include{chapters/12j-lists}
\include{chapters/12k-tables}
\include{chapters/12l-trees}
\include{chapters/12m-editors}
\include{chapters/12n-buttons}
\include{chapters/12o-progress}
\include{chapters/12p-menus}
\include{chapters/12q-dialogs}
\include{chapters/12r-toasts}
\include{chapters/12s-splits}
\include{chapters/12t-tabs}
\include{chapters/12u-statusbar}
\include{chapters/12v-pubsub}
\include{chapters/12w-fileio}
\include{chapters/12x-keymaps}
\include{chapters/12y-theming}
\include{chapters/12z-persistence}
\include{chapters/12aa-logging}
\include{chapters/12bb-plugins}
\include{chapters/12cc-pty}
\include{chapters/12dd-networking}
\include{chapters/12ee-a11y2}
\include{chapters/12ff-i18n2}
\include{chapters/12gg-security2}
\include{chapters/12hh-crossplat2}
\include{chapters/12ii-cmake}
\include{chapters/12jj-profile}
\include{chapters/12kk-memory}
\include{chapters/12ll-testing2}
\include{chapters/12mm-customwidget}
\include{chapters/12nn-curves}
\include{chapters/12oo-fsm}
\include{chapters/12pp-dnd}
\include{chapters/12qq-search}
\include{chapters/12rr-autocomplete}
\include{chapters/12ss-charts}
\include{chapters/12tt-pickers}
\include{chapters/12uu-diffview}
\include{chapters/12vv-markdown}
\include{chapters/12ww-syntax}
\include{chapters/12xx-undostate}
\include{chapters/12yy-hotreload}
\include{chapters/12zz-help}
\include{chapters/12ax-spinners}
\include{chapters/13a-offline}
\include{chapters/13b-selection}
\include{chapters/13c-signaldbg}
\include{chapters/13d-mvp}
\include{chapters/13e-lazy}
\include{chapters/13f-errorboundary2}
\include{chapters/13g-slots}
\include{chapters/13h-flexbox2}
\include{chapters/13i-grid}
\include{chapters/13j-optcases}
\include{chapters/13k-concurrency}
\include{chapters/13l-limits}
\include{chapters/13m-chatclient}
\include{chapters/13n-procmon}
\include{chapters/13o-filemgr}
\include{chapters/13p-musicplayer}
\include{chapters/13q-migrating}
\include{chapters/13r-distrib}
\include{chapters/13s-contributing}
\include{chapters/13t-catalogue1}
\include{chapters/13u-catalogue2}
\include{chapters/13v-catalogue3}
\include{chapters/13w-catalogue4}
\include{chapters/13x-catalogue5}
\include{chapters/13y-glossary2}
\include{chapters/13z-antipatterns}
\include{chapters/14a-futuregraphics}
\include{chapters/14b-futureinput}
\include{chapters/14c-futurelib}
\include{chapters/14d-web}
\include{chapters/14e-commonbugs}
\include{chapters/14f-recipes}
\include{chapters/14g-perfbudget}
\include{chapters/14h-loopannotated}
\include{chapters/14i-readingsource}
\include{chapters/14j-nextsteps}
\include{chapters/14k-colocation}
\include{chapters/14l-dataflow}
\include{chapters/14m-derived}
\include{chapters/14n-commands}
\include{chapters/14o-providers}
\include{chapters/14p-portals}
\include{chapters/14q-renderbudget}
\include{chapters/14r-apidesign}
\include{chapters/14s-docs}
\include{chapters/14t-v1}
\include{chapters/14u-conclusion}
\include{chapters/14v-uistate}
\include{chapters/14w-optimistic}
\include{chapters/14x-lazystate}
\include{chapters/14y-vscroll}
\include{chapters/14z-keydisc}
\include{chapters/15a-degradation}
\include{chapters/15b-progdisc}
\include{chapters/15c-empty}
\include{chapters/15d-formergo}
\include{chapters/15e-undobranches}
\include{chapters/15f-animchain}
\include{chapters/15g-timers}
\include{chapters/15h-flags}
\include{chapters/15i-exit}
\include{chapters/15j-afterword}
\include{chapters/15k-ansiref}
\include{chapters/15l-keyref}
\include{chapters/15m-signalref}
\include{chapters/15n-elementref}
\include{chapters/15o-troubleshoot}
\include{chapters/15p-further}
\include{chapters/15q-capref}
\include{chapters/15r-escglossary}
\include{chapters/15s-patternindex}
\include{chapters/15t-buildrecipes}
\include{chapters/15u-deployments}
\include{chapters/15v-partingchecklist}
\include{chapters/15w-refman}
\include{chapters/15x-refman-program}
\include{chapters/15y-refman-run}
\include{chapters/15z-refman-text}
\include{chapters/15za-refman-rowcol}
\include{chapters/15zb-refman-padding}
\include{chapters/15zc-refman-style}
\include{chapters/15zd-refman-align}
\include{chapters/15ze-refman-sub}
\include{chapters/15zf-refman-keymap}
\include{chapters/15zg-refman-timers}
\include{chapters/15zh-refman-cmd}
\include{chapters/15zi-refman-task}
\include{chapters/15zj-refman-mouse}
\include{chapters/15zk-refman-when}
\include{chapters/15zl-refman-list}
\include{chapters/15zm-refman-input}
\include{chapters/15zn-refman-progress}
\include{chapters/15zo-refman-spinner}
\include{chapters/15zp-refman-table}
\include{chapters/15zq-refman-tabs}
\include{chapters/15zr-refman-modal}
\include{chapters/15zs-refman-scroll}
\include{chapters/15zt-refman-print}
\include{chapters/15zu-refman-live}
\include{chapters/15zv-refman-focus}
\include{chapters/15zw-refman-signals}
\include{chapters/15zx-refman-resize}
\include{chapters/15zy-refman-errors}
\include{chapters/15zz-refman-exit}
\include{chapters/16aa-refman-lifecycle}
\include{chapters/16ab-refman-compose}
\include{chapters/16ac-refman-test}
\include{chapters/16ad-refman-cmake}
\include{chapters/16ae-refman-apps}
\include{chapters/16af-refman-cookbook}
\include{chapters/16ag-refman-enums}
\include{chapters/13-events}
\include{chapters/14-cmdsub}
\include{chapters/15-widgets}
\include{chapters/16-build}

\backmatter
\include{chapters/99-colophon}

\end{document}
